### 2014-12-03 docs/ms.Rmd
Tighten intro some more, hack back methods bit

# Reproducible research in biological colour science

@@

Keywords: #colsci, spectrometry, photography, visual ecology, pigment, structural colour, visual modeling, methodology

@@

Studies of biological colouration have produced insights into fundamental evolutionary and ecological processes, and research into colour traits is a rapidly growing, interdisciplinary field []. The resurgent interest in biological colouration has in large part been driven by the increased availability of key technologies, including spectrometry and photography, and concurrent developments in methods for analysing colour data, such as visual models []. While such advances are extremely positive for the field, the rapid rate of progression means that a deep understanding of common methods is increasingly beyond the grasp of the average researcher. Indeed, the subtle complexity of these techniques [e.g. the 3-7 parameters that define a typical visual model; ] is often overlooked by empiricists, which leads to critical methodological information often not being reported (Fig. 1); thus limiting a study's utility.

Replication and transparency lie at the heart of the scientific enterprise. True replication of individual studies based on independantly collected data is the gold standard in cumulative science. This may often be unfeasable however, espescially in field-based research, which includes a great deal of research on colour-traits []. In contrast, reproducability - the re-analysis of a study's data by independant researchers - has been presented as a readily attainable minimum standard [@peng_reproducible_2006; @peng_reproducible_2011]. A study may be considered truly reproducible when it satifies three broad criteria: (i) methods are reported completely, (ii) data are publicly archived and available, and (iii) analysis code is made publicly available, such that the chain of modification of raw data is documented and preserved. While completely reproducible research [e.g. @fitzjohn_much_2014] is a laudable goal, the considerable demands on researchers means that it will often, in practice, be unattanable. Nevertheless, even partial reproducibility is of tremendous value as it allows greater comparability between studies, and may serve as a launching point for future research.

Our aim here to explore current levels of reproducibility in studies of biological colouration, and to outline ways in which they may be improved in line with points (i) - (iii) above. We begin by briefly outlining some of the most common methods for studying biological colouration. We then present a set of guidelines for the reporting of methods for the capture and analysis of colour data, and we assess a subset of the literature against these criteria to assess the current state of methods reporting in colour science. We also quantify the availability of publicly archived data and code, and suggest some tools...

@@

todo: hack this back

Generations of biologists have endeavored to explain the mechanisms and functions of animal and plant colouration [@thayer_concealing_1909; @poulton_colours_1890; @wallace_natural_1891], and identifying best practices in measuring colour has been a great challenge. The application of spectrometry to questions of biological colouration revolutionized the field [@dyck_determination_1966]. Before its widespread use, researchers assigned categorical colour descriptions, arbitrary rankings or scores [e.g. @aldrich_ecogeographic_1991; @bailey_latitudinal_1978; @wheelwright_colors_1985] or matched paint swatches or munsell chip ratings based on their own visual assessment [e.g. @burtt_analysis_1986; @key_temperature_1954]. Although no doubt applied carefully, these techniques limited progress. For example, many researchers designed colour assessment methods specifically for their projects, rendering data incomparable across studies [e.g. @aldrich_ecogeographic_1991; @bailey_latitudinal_1978]. This is especially problematic when different measurement techniques are used on the same species [e.g. house finch _Carpodacus mexicanus_; @aldrich_origin_1978; @hill_female_1990]. Likewise, because human vision is inherently different to most other animals [@bennett_sexual_1994; @douglas_spectral_2014], its use has lead to erroneous conclusions [@bennett_sexual_1994; @endler_measurement_1990].

Measuring reflectance spectra via spectrometry is now a reasonably inexpensive and convenient technique. Spectrometry allows the capture of quantitative and arguably repeatable data and has been rapidly taken up by researchers the field. However one drawback to this methodology, is the ease with which erroneous data can be collected if attention is not paid to finer details of spectrum capture. The angle from which spectra are taken [@santos_strong_2007], consistent calibration of light and dark standards, the distance between light source and surface, and the size, shape, glossiness or iridescence of the specimen can all contribute to variation in data and potentially bias results [@andersson_quantifying_2006; @kemp_ornamental_2008; @meadows_quantifying_2011; @schaefer_birds_2008]. Experience, guidance, consistency and objectivity (e.g. double-blind experimental design) are necessary to ensure data are repeatable and representative of the subject. Crucially, to ensure repeatability across studies, the conditions under which data were collected must be reported exactly.

Complementary to spectrometry, digital photography is increasingly being used to quantify biological colouration [@stevens_using_2007]. High-resolution cameras are inexpensive,  and allow instantaneous, repeatable sampling of multiple colour patches [@mckay_use_2013]. The main challenge for gathering colour data from photographs involve ensuring standardization, such as linearization and consistent lighting [@stevens_using_2007]. Also, most digital cameras used for the measurement of biological colouration are consumer products designed for the human visual system. Thus hardware modifications for ultraviolet and infrared imaging are generally necessary [e.g. @stevens_colour_2013].

Expansion in the availability of objective methods for the measurement of colour has been matched by advances in theory and analysis. In particular, the development of sensory models has produced significant insights into how colourful stimuli are perceived [@chittka_colour_1992; @endler_comparing_2005; @vorobyev_receptor_1998]. Sensory models typically attempt to describe the subjective perception of colour or brightness information as a function of an object’s reflectance, the ambient illumination, and a receiver’s sensory system [e.g. @endler_comparing_2005; @vorobyev_receptor_1998]. These models enable researchers to move beyond purely quantitative comparisons of reflectance spectra and adopt biologically relevant perspectives of signal receivers to formulate and test hypotheses.

Although relatively easy to implement, and potentially powerful, sensory models are built on multiple assumptions about the way in which stimuli are visually and perceptually processed and these can dramatically shape the results of a given analysis [@lind_avian_2009; @pike_preserving_2012]. For example, Vorobyev et al.’s [-@vorobyev_receptor_1998; -@vorobyev_colour_2001] widely used receptor-noise model requires researchers to specify the form of receptor quantum catches (i.e. raw versus log-transformed), the use of any colour constancy mechanism (e.g. von Kries), as well as the photoreceptor signal-to-noise ratios, their relative densities, and the type of noise predominating (i.e. neural and/or receptor noise). These data are only available for a small number of species, such as humans and honeybees, and so assumptions and extrapolations are necessary when models are applied to other species. Unless the biological rationale underlying every parameter choice is clearly understood, justified, and reported, such analyses may become ineffective, especially when comparing results between studies [@pike_preserving_2012].

The potential for subjectivity in colour measurement and the mathematical opacity of visual models for the average biologist means that, as in any field, is it critical that methods are reported rigorously. Methods reporting should ideally have well defined standards such that sufficient detail is provided to allow comprehensive peer review, precise replication, cross-study comparison and data sharing [@alsheikh_public_2011; @sandve_ten_2013]. The aims of this paper were to (a) review methods reporting in papers published on colour in 2013 from 22 top tier journals and (b) provide a set of recommendations for best-practice in reporting on the most common methods. Our review specifically focuses on papers that quantify the external coloration of a plant or animal. These may be purely descriptive quantifications of color, or they may be incorporated into models of animal vision. Our hope is that this paper will provide a useful set of guidelines to improve the reproducibility and transparency of studies on colour.

@@

To determine the current state of reproducibility in the field we searched for papers from 2013 in 22 leading journals (_American Journal of Botany_, _The American Naturalist_, _Animal Behaviour_, _Behavioral Ecology_, _Behavioral Ecology and Sociobiology_, _Biological Journal of the Linnean Society_, _Biology Letters_, _Current Zoology_, _Ecology_, _Ecology and Evolution_, _Ecology Letters_, _Ethology_, _Evolution_, _Functional Ecology_, _Journal of Ecology_, _The Journal of Evolutionary Biology_, _The Journal of Experimental Biology_, _Naturwissenschaften_, _New Phytologist_, _Oikos_, _PLoS ONE_, _Proceedings of the Royal Society B: Biological Sciences_). On each of the journals' homepages, we used the Boolean terms 'colour*', 'color*' or 'spectra*' to search the title or abstract. We read the titles and abstracts of each paper and included only those papers that used either a spectrometer or camera to quantify colouration. We excluded papers review papers, methodological papers, papers quantifying colour patterns rather than the chromatic properties of a colour patch, and papers taking microspectrophotometric measurments of retinal absorbance. To review the large number of papers returned from our search, journals were haphazardly divided up between the authors. In order to reduce the risk of observer bias in our assesment, each paper was read and reassessed by three authors. Any discrepancies between assessor scores were discussed and resolved prior to analysis. We then tabulated what details of the methods were reported against our criteria. The data along with analysis script can be found at: https://github.com/daniel1noble/colsci_rep (TODO: GIVE IT A DOI?). We have kept the papers used in our dataset annoymous as our aim is to make a general statement about methods reporting in the field.

@@

While all criteria are important, some deserve special attention. In particular, we found that the types of light sources, numbers of pixels or spectra averaged, and types of standards used (i.e. white and black reference samples) are under-reported, each of which is critical to repeatability and cross-comparison. Similarly, the distance and angle of the optical probe relative to the measurment surface and collector should be standard reporting for all studies using spectrophometry, as variation in these factors can lead to dramatically different results in some cases (REFS). Visual models were regularly used to analyse spectral data, however, there were many important aspects of the analysis that were seldom reported. These included the type of quantum catch used, whether photoreceptor adaptation was modelled and the type of noise used in receptor-noise models. This is problematic because... Also, photoreceptor densities should be reported where possible even if using standard avian photorecptor densities because.... These all have important consequences on downstream analysis and the reproducibility of colour signal quantification [e.g. @kemp_ornamental_2008; @santos_comparison_2007]. We also recommend the presentation of spectral reflectance curves where possible. These figures allow for the rapid assesment of the nature and quality of the data, independent of downstream processing (if the conditions under which they were captured is comprehensively reported. We found that only `r round(propSpec_fin["prop_refl_fig"], digits = 2)`% of studies using spectral reflectance data had a figure showing these curves. When possible, we also recommend that authors justify data capture and analytical approaches - something few studies we investigated explicitly outlined.

@@

We believe that part of the issue at hand is simply that many researchers seldom have clear guidelines to follow when reporting methods, and may overlook outlining important aspects of methodology and analysis. We hope our critical evaluation of recent literature has highlighted a need for more careful description of how colour signals have been measured and analyzed. More importantly, we hope our guidelines for methods reporting helps establish a standard in this area, to continue the advance of an exciting era in colour research.

>>>>

# Reproducible research in the study of biological colouration

@@

Keywords: #colsci, spectrometry, photography, visual ecology, pigment, structural colour, visual modelling, methodology

@@

Studies of biological colouration have generated insights into fundamental evolutionary and ecological processes, and research into colour traits is a rapidly growing field []. The resurgent interest in biological colouration has in large part been driven by the increased availability of key technologies, including spectrometry and photography, and concurrent developments in methods for analysing colour data, such as visual models [@endler_comparing_2005; @stevens_using_2007; @kelber_animal_2003]. While such advances are positive for the field, the rapid rate of progression means that a deep understanding of common methods is increasingly beyond the grasp of the average researcher. As a consequence, the subtle complexity of many techniques [e.g. the 3-7 parameters that define a typical visual model; @endler_comparing_2005; @vorobyev_receptor_1998] is often overlooked by empiricists, which leads to critical methodological information often not being reported (Fig. 1).

Replication and transparency lie at the heart of the scientific enterprise. The full replication of individual studies based on independently collected data is the gold standard in cumulative science. This may often be unfeasible however, especially in field-based research, which includes a great deal of research on colour-traits []. In contrast, reproducibility - the re-analysis of a study's data by independent researchers - has been presented as a readily attainable alternative [@peng_reproducible_2006; @peng_reproducible_2011]. A study may be considered truly reproducible when it satisfies three broad criteria: (i) methods are reported completely, (ii) data are publicly archived and available, and (iii) analysis code is made publicly available, such that the chain of modification of raw data is documented and preserved. While completely reproducible research [e.g. @fitzjohn_much_2014] is a laudable goal, the considerable demands it imposes on researchers means that it will often, in practice, be unattainable. Nevertheless, even partial reproducibility is of tremendous value both to individual researchers and the broader community as it ensures greater comparability between studies, and may provide a launching point for future research.

Our aim here to explore current levels of reproducibility in studies of biological colouration, and to outline simple ways in which they may be improved in line with points (i) - (iii) above. We begin by briefly outlining some of the most common methods for studying biological colouration. We then present a set of guidelines for the reporting of methods, and we assess a subset of the literature against these criteria to assess the current state of methods reporting in colour science. We also analyse the availability of publicly archived data and code, and suggest some tools that may be useful for increasing reproducibility.

@@

todo: hack this back, make it more relevant to reproducibility

Generations of biologists have endeavoured to explain the mechanisms and functions of animal and plant colouration [@thayer_concealing_1909; @poulton_colours_1890; @wallace_natural_1891], and identifying best practices in measuring colour has been a great challenge. The application of spectrometry to questions of biological colouration revolutionized the field [@dyck_determination_1966]. Now relatively inexpensive and convenient, spectrometry allows the capture of quantitative and arguably repeatable data and has been rapidly taken up by researchers the field. However one drawback to this methodology, is the ease with which erroneous data can be collected if attention is not paid to finer details of spectrum capture. The angle from which spectra are taken [@santos_strong_2007], consistent calibration of light and dark standards, the distance between light source and surface, and the size, shape, glossiness or iridescence of the specimen can all contribute to variation in data and potentially bias results [@andersson_quantifying_2006; @kemp_ornamental_2008; @meadows_quantifying_2011; @schaefer_birds_2008]. Experience, guidance, consistency and objectivity (e.g. double-blind experimental design) are necessary to ensure data are repeatable and representative of the subject. Crucially, to ensure repeatability across studies, the conditions under which data were collected must be reported exactly.

Complementary to spectrometry, digital photography is increasingly being used to quantify biological colouration [@stevens_using_2007]. High-resolution cameras are inexpensive,  and allow instantaneous, repeatable sampling of multiple colour patches [@mckay_use_2013]. The main challenge for gathering colour data from photographs involve ensuring standardization, such as linearisation and consistent lighting [@stevens_using_2007]. Also, most digital cameras used for the measurement of biological colouration are consumer products designed for the human visual system. Thus hardware modifications for ultraviolet and infra-red imaging are generally necessary [e.g. @stevens_colour_2013].

Expansion in the availability of objective methods for the measurement of colour has been matched by advances in theory and analysis. In particular, the development of visual models has produced significant insights into the early visual processing of colourful stimuli [@chittka_colour_1992; @endler_comparing_2005; @vorobyev_receptor_1998] and, in a few cases, how they may be perceived [@kelber_spectral_2010]. Visual models typically attempt to describe the subjective perception of colour information as a function of an object’s reflectance, the ambient illumination, and a receiver’s sensory system [e.g. @endler_comparing_2005; @vorobyev_receptor_1998]. These models enable researchers to move beyond purely quantitative comparisons of reflectance spectra and adopt more biologically relevant perspectives to refine and test hypotheses.

Although relatively easy to implement, and potentially powerful, sensory models are built on multiple assumptions about the way in which stimuli are visually processed, that can dramatically shape the results of a given analysis [@lind_avian_2009; @pike_preserving_2012]. For example, Vorobyev et al.’s [-@vorobyev_receptor_1998; -@vorobyev_colour_2001] widely used receptor-noise model requires researchers to specify the form of receptor quantum catches (i.e. raw versus log-transformed), the use of any colour constancy mechanism (e.g. von Kries), as well as the photoreceptor signal-to-noise ratios, their relative densities, and the type of noise predominating (i.e. neural and/or receptor noise). These data are only available for a small number of species, such as humans and honeybees, and so assumptions and extrapolations are necessary when models are applied to other species. Unless the biological rationale underlying every parameter choice is clearly understood, justified, and reported, such analyses may become ineffective, especially when comparing results between studies [@pike_preserving_2012].

The potential for subjectivity in colour measurement and the mathematical opacity of visual models for the average biologist means that, as in any field, is it critical that methods are reported rigorously. Methods reporting should ideally have well defined standards such that sufficient detail is provided to allow comprehensive peer review, precise replication, cross-study comparison and data sharing [@alsheikh_public_2011; @sandve_ten_2013]....

@@

To assess the current state of reproducibility in the field we searched papers from 2013 in 22 leading journals (_American Journal of Botany_, _The American Naturalist_, _Animal Behaviour_, _Behavioral Ecology_, _Behavioral Ecology and Sociobiology_, _Biological Journal of the Linnean Society_, _Biology Letters_, _Current Zoology_, _Ecology_, _Ecology and Evolution_, _Ecology Letters_, _Ethology_, _Evolution_, _Functional Ecology_, _Journal of Ecology_, _The Journal of Evolutionary Biology_, _The Journal of Experimental Biology_, _Naturwissenschaften_, _New Phytologist_, _Oikos_, _PLoS ONE_, _Proceedings of the Royal Society B: Biological Sciences_). On each of the journals' homepages, we used the Boolean terms 'colour*', 'color*' or 'spectra*' to search the title or abstract. We read the titles and abstracts of each paper and included only those papers that used either a spectrometer or camera to quantify colouration. We excluded papers review papers, methodological papers, papers quantifying colour patterns rather than the chromatic properties of a colour patch, and papers taking micro-spectrometric measurements of retinal absorbance. To review the large number of papers returned from our search, journals were haphazardly divided up between the authors. In order to reduce the risk of observer bias in our assesment, each paper was read and reassessed by three authors. Any discrepancies between assessor scores were discussed and resolved prior to analysis. We then tabulated what details of the methods were reported against our criteria. The data along with analysis script can be found at: https://github.com/daniel1noble/colsci_rep (TODO: GIVE IT A DOI?). We have kept the papers used in our dataset anonymous as our aim is to make a general statement about methods reporting in the field.

@@

While all criteria are important, some deserve special attention. In particular, we found that the types of light sources, numbers of pixels or spectra averaged, and types of standards used (i.e. white and black reference samples) are under-reported, each of which is critical to repeatability and cross-comparison. Similarly, the distance and angle of the optical probe relative to the measurement surface and collector should be standard reporting for all studies using spectrometry, as variation in these factors can lead to dramatically different results in some cases (REFS). Visual models were regularly used to analyse spectral data, however, there were many important aspects of the analysis that were seldom reported. These included the type of quantum catch used, whether photoreceptor adaptation was modelled and the type of noise used in receptor-noise models. This is problematic because... Also, photoreceptor densities should be reported where possible even if using standard avian photoreceptor densities because.... These all have important consequences on downstream analysis and the reproducibility of colour signal quantification [e.g. @kemp_ornamental_2008; @santos_comparison_2007]. We also recommend the presentation of spectral reflectance curves where possible. These figures allow for the rapid assessment of the nature and quality of the data, independent of downstream processing (if the conditions under which they were captured is comprehensively reported. We found that only `r round(propSpec_fin["prop_refl_fig"], digits = 2)`% of studies using spectral reflectance data had a figure showing these curves. When possible, we also recommend that authors justify data capture and analytical approaches - something few studies we investigated explicitly outlined.

@@

We believe that part of the issue at hand is simply that many researchers seldom have clear guidelines to follow when reporting methods, and may overlook outlining important aspects of methodology and analysis. We hope our critical evaluation of recent literature has highlighted a need for more careful description of how colour signals have been measured and analysed. More importantly, we hope our guidelines for methods reporting helps establish a standard in this area, to continue the advance of an exciting era in colour research.
